---
title: "606 Lab 9"
author: "Lu Beyer"
output: html_document
---

```{r}
library(tidyverse)
library(openintro)
library(DATA606)
library(GGally)
```

```{r}
x <- evals
```


```{r}
glimpse(evals)
```
```{r}
?evals
```

Is this an observational study or an experiment? The original research question posed in the paper is whether beauty leads directly to the differences in course evaluations. Given the study design, is it possible to answer this question as it is phrased? If not, rephrase the question.

This is an an observational study. I don't know that it's possible to accurately answer this question, expecially given that the 463 students evaluating the professors  are unlikely to all agree with the six students assessing the professor's physical appearance. 

?????


Exercise 2
Describe the distribution of score. Is the distribution skewed? What does that tell you about how students rate courses? Is this what you expected to see? Why, or why not?

```{r}
ggplot(data = x, aes(x = score)) +
  geom_histogram()
```

```{r}
hist(evals$score)
```



The data is left skewed, which I would expect to some degree just from experience o being in school and talking to my peers who more often either expressed positive feelings about the majority of their professors than negative.  I would have expected a slightly more normal distribution with data that is more symmetric, but I also feel my experience with peers, students either liked or didn't like their professors rather than feeling neutral.  


Exercise 3
Excluding score, select two other variables and describe their relationship with each other using an appropriate visualization.

```{r}
y <- x %>% 
  beautyavg <- mean()
```


```{r}
ggplot(data = x, aes(x = gender, y = cls_students)) +
  geom_bar(stat = "identity", position = "dodge")
```









