---
title: "606 Final"
author: "Lu Beyer"
date: "4/27/2024"
output: pdf_document
---
# Abstract

# Introduction

# Data	

```{r}
library(tidyverse)
library(infer)

data <- read.csv("Buf_Roch_2022.csv") %>% 
  rename(Category = 1,
         Buf_Population = 2,
         Buf_Population_Disability = 3,
         Buf_Disability_Percentage = 4,
         Roch_Population = 5,
         Roch_Population_Disability = 6,
         Roch_Disability_Percentage = 7) %>% 
  mutate(across(where(is.character), ~str_replace_all(., "Â", ""))) %>%
  mutate(Category = str_trim(Category)) %>% 
  mutate(across(where(is.character), ~str_replace_all(., "^N$", as.character(NA)))) %>% 
  mutate(across(c(Buf_Population, Buf_Population_Disability), ~as.numeric(str_replace_all(., ",", "")))) %>% 
  mutate(across(c(Roch_Population, Roch_Population_Disability), ~as.numeric(str_replace_all(., ",", "")))) %>% 
  mutate(Buf_Disability_Percentage = as.numeric(str_remove(Buf_Disability_Percentage, "%")) / 100) %>% 
  mutate(Roch_Disability_Percentage = as.numeric(str_remove(Roch_Disability_Percentage, "%")) / 100) %>% 
  mutate(Buf_Population_Nondisabled = Buf_Population - Buf_Population_Disability,
         Roch_Population_Nondisabled = Roch_Population - Roch_Population_Disability) %>% 
  select(Category, Buf_Population, Buf_Population_Disability, Buf_Population_Nondisabled, Buf_Disability_Percentage,
         Roch_Population, Roch_Population_Disability, Roch_Population_Nondisabled, Roch_Disability_Percentage)



sex <- data %>% 
  slice(3:4)

race <- data %>% 
  slice(6:14)

age <- data %>% 
  slice(16:21)

general <- data %>% 
  slice(1, 3:4, 6:14, 16:21) %>% 
  drop_na()

head(general, 5)
```

# Summary Statistics

### Disabilty Rates by Age
```{r}
cat <- c("Under 5 years", "5 to 17 years", "18 to 34 years", "35 to 64 years", "65 to 74 years", "75 years and over")

age_long <- age %>%
  pivot_longer(
    cols = c(Buf_Disability_Percentage, Roch_Disability_Percentage),
    names_to = "City",
    values_to = "Disability_Percentage"
  ) %>%
  mutate(Category = factor(Category, levels = rev(cat))) 

ggplot(data = age_long, aes(x = Category, y = Disability_Percentage, fill = City)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip()
```

### Disability Rates by Sex
```{r}
sex_long <- sex %>%
  pivot_longer(
    cols = c(Buf_Disability_Percentage, Roch_Disability_Percentage),
    names_to = "City",
    values_to = "Disability_Percentage"
  )

ggplot(data = sex_long, aes(x = Category, y = Disability_Percentage, fill = City)) +
  geom_bar(stat = "identity", position = "dodge")
```
### Disabiliy Rates by Race
```{r}
race_long <- race %>%
  pivot_longer(
    cols = c(Buf_Disability_Percentage, Roch_Disability_Percentage),
    names_to = "City",
    values_to = "Disability_Percentage"
  )

ggplot(data = race_long, aes(x = Category, y = Disability_Percentage, fill = City)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip()
```


# Chi-Square Tests

### General Population
```{r}
general_table <- matrix(c(general$Buf_Population_Disability[1], general$Buf_Population_Nondisabled[1],
                        general$Roch_Population_Disability[1], general$Roch_Population_Nondisabled[1]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_general <- chisq.test(general_table)
print(chisq_result_general)

barplot(general_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(general_table) * 1.4),
        legend = rownames(general_table),
        main = "Count of General Population Difference in Disability Status")

print(chisq_result_general$residuals)
```

### Population - Under 5
```{r}
#Under 5
age_table <- matrix(c(age$Buf_Population_Disability[1], age$Buf_Population_Nondisabled[1],
                        age$Roch_Population_Disability[1], age$Roch_Population_Nondisabled[1]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_age <- chisq.test(age_table)
print(chisq_result_age)

barplot(age_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(age_table) * 1.4),
        legend = rownames(age_table),
        main = "Difference in Disability Status, Age Under 5 Years")


chisq_result_age$stdres
```

### Population - Between 5 to 17 Years
```{r}
#5 to 17 Years
age_table <- matrix(c(age$Buf_Population_Disability[2], age$Buf_Population_Nondisabled[2],
                        age$Roch_Population_Disability[2], age$Roch_Population_Nondisabled[2]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_age <- chisq.test(age_table)
print(chisq_result_age)

barplot(age_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(age_table) * 1.4),
        legend = rownames(age_table),
        main = "Difference in Disability Status, Age 5 to 17 Years")


chisq_result_age$stdres
```

### Population - Between 18 and 34 Years
```{r}
#18 to 34 Years
age_table <- matrix(c(age$Buf_Population_Disability[3], age$Buf_Population_Nondisabled[3],
                        age$Roch_Population_Disability[3], age$Roch_Population_Nondisabled[3]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_age <- chisq.test(age_table)
print(chisq_result_age)

barplot(age_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(age_table) * 1.4),
        legend = rownames(age_table),
        main = "Difference in Disability Status, Age 5 to 17 Years")


chisq_result_age$stdres
```

### Population - Two or More Races
```{r}
#Two or More Races
race_table <- matrix(c(race$Buf_Population_Disability[7], race$Buf_Population_Nondisabled[7],
                        race$Roch_Population_Disability[7], race$Roch_Population_Nondisabled[7]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(race_table) * 1.4),
        legend = rownames(race_table),
        main = "Difference in Disability Status, Two or More Races")


chisq_result_race$stdres
```


### Population - White Alone
```{r}
#White
race_table <- matrix(c(race$Buf_Population_Disability[1], race$Buf_Population_Nondisabled[1],
                        race$Roch_Population_Disability[1], race$Roch_Population_Nondisabled[1]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(race_table) * 1.4),
        legend = rownames(race_table),
        main = "Difference in Disability Status, White Alone")


chisq_result_race$stdres
```

### Population - White Alone, Not Hispanic or Latino
```{r}
#White Alone - not Hispanic or Latino
race_table <- matrix(c(race$Buf_Population_Disability[8], race$Buf_Population_Nondisabled[8],
                        race$Roch_Population_Disability[8], race$Roch_Population_Nondisabled[8]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(race_table) * 1.4),
        legend = rownames(race_table),
        main = "Difference in Disability Status, White Alone - Not Hispanic or Latino")


chisq_result_race$stdres
```

### Population - Hispanic or Latino
```{r}
#Hispanic or Latino
race_table <- matrix(c(race$Buf_Population_Disability[9], race$Buf_Population_Nondisabled[9],
                        race$Roch_Population_Disability[9], race$Roch_Population_Nondisabled[9]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(race_table) * 1.4),
        legend = rownames(race_table),
        main = "Difference in Disability Status, Hispanic or Latino (of any race)")


chisq_result_race$stdres
```

### Population - American Indian and Alaska Native Alone

```{r}
#American Indian and Alaska Native alone

race_table <- matrix(c(race$Buf_Population_Disability[3], race$Buf_Population_Nondisabled[3],
                        race$Roch_Population_Disability[3], race$Roch_Population_Nondisabled[3]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        legend = rownames(race_table),
        ylim = c(0, max(race_table) * 1.4),
        main = "Difference in Disability Status, American Indian and Alaska Native")


chisq_result_race$stdres
```


### Population - Men
```{r}
males_table <- matrix(c(sex$Buf_Population_Disability[1], sex$Buf_Population_Nondisabled[1],
                        sex$Roch_Population_Disability[1], sex$Roch_Population_Nondisabled[1]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_males <- chisq.test(males_table)
print(chisq_result_males)


barplot(males_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        legend = rownames(males_table),
        ylim = c(0, max(males_table) * 1.4),
        main = "Difference in Disability Status, Men")


chisq_result_males$stdres

```

### Population - Women
```{r}
females_table <- matrix(c(sex$Buf_Population_Disability[2], sex$Buf_Population_Nondisabled[2],
                        sex$Roch_Population_Disability[2], sex$Roch_Population_Nondisabled[2]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_females <- chisq.test(females_table)
print(chisq_result_females)

barplot(females_table,
        beside = TRUE, 
        col = c("lightblue", "lightpink"),
        legend = rownames(females_table),
        ylim = c(0, max(females_table) * 1.5),
        main = "Difference in Disability Status, Women")


chisq_result_females$stdres

```




# Conclusion








