---
title: "606 Final"
author: "Lu Beyer"
date: "4/27/2024"
output: pdf_document
---
# Abstract

# Introduction



# Research Questions
Null Hypothesis:  There is no statistically significant difference in disability rate between Buffalo and Rochester NY
Alternate Hypothesis:  There is a significant statistical difference in disability rate between Buffalo and Rochester NY

# Data	
Data was pulled from US Census, through the 2022 American Community Survey - https://data.census.gov/table/ACSST1Y2022.S1810?q=disability&g=310XX00US15380,40380&moe=false&tp=false

We are looking at Buffalo and Rochester Metro Areas, rather than strictly city limits, as they are closer in terms of total population for comparison.

```{r}
library(tidyverse)
library(infer)

data <- read.csv("Buf_Roch_2022.csv") %>% 
  rename(Category = 1,
         Buf_Population = 2,
         Buf_Population_Disability = 3,
         Buf_Disability_Percentage = 4,
         Roch_Population = 5,
         Roch_Population_Disability = 6,
         Roch_Disability_Percentage = 7) %>% 
  mutate(across(where(is.character), ~str_replace_all(., "Â", ""))) %>%
  mutate(Category = str_trim(Category)) %>% 
  mutate(across(where(is.character), ~str_replace_all(., "^N$", as.character(NA)))) %>% 
  mutate(across(c(Buf_Population, Buf_Population_Disability), ~as.numeric(str_replace_all(., ",", "")))) %>% 
  mutate(across(c(Roch_Population, Roch_Population_Disability), ~as.numeric(str_replace_all(., ",", "")))) %>% 
  mutate(Buf_Disability_Percentage = as.numeric(str_remove(Buf_Disability_Percentage, "%")) / 100) %>% 
  mutate(Roch_Disability_Percentage = as.numeric(str_remove(Roch_Disability_Percentage, "%")) / 100) %>% 
  mutate(Buf_Population_Nondisabled = Buf_Population - Buf_Population_Disability,
         Roch_Population_Nondisabled = Roch_Population - Roch_Population_Disability) %>% 
  select(Category, Buf_Population, Buf_Population_Disability, Buf_Population_Nondisabled, Buf_Disability_Percentage,
         Roch_Population, Roch_Population_Disability, Roch_Population_Nondisabled, Roch_Disability_Percentage)



sex <- data %>% 
  slice(3:4)

race <- data %>% 
  slice(6:14)

age <- data %>% 
  slice(16:21)

general <- data %>% 
  slice(1, 3:4, 6:14, 16:21) %>% 
  drop_na()

head(general, 5)
```

# Summary Statistics

### Disabilty Rates by Age
```{r}
cat <- c("Under 5 years", "5 to 17 years", "18 to 34 years", "35 to 64 years", "65 to 74 years", "75 years and over")

age_long <- age %>%
  pivot_longer(
    cols = c(Buf_Disability_Percentage, Roch_Disability_Percentage),
    names_to = "City",
    values_to = "Disability_Percentage"
  ) %>%
  mutate(Category = factor(Category, levels = rev(cat))) 

ggplot(data = age_long, aes(x = Category, y = Disability_Percentage, fill = City)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip()
```

### Disability Rates by Sex
```{r}
sex_long <- sex %>%
  pivot_longer(
    cols = c(Buf_Disability_Percentage, Roch_Disability_Percentage),
    names_to = "City",
    values_to = "Disability_Percentage"
  )

ggplot(data = sex_long, aes(x = Category, y = Disability_Percentage, fill = City)) +
  geom_bar(stat = "identity", position = "dodge")
```
### Disabiliy Rates by Race
```{r}
race_long <- race %>%
  pivot_longer(
    cols = c(Buf_Disability_Percentage, Roch_Disability_Percentage),
    names_to = "City",
    values_to = "Disability_Percentage"
  )

ggplot(data = race_long, aes(x = Category, y = Disability_Percentage, fill = City)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip()
```

Looking at the summary statistic charts, I’ve identified the following demographics as the most likely to have a statistically significant difference in disability rates, due to their notably large differences in rates

Age:
0 - 5
5 - 17

Race:
Two or more races
Some other race
Hispanic
American Indian and Alaska Native


# Chi-Square Tests

### General Population
```{r}
general_table <- matrix(c(general$Buf_Population_Disability[1], general$Buf_Population_Nondisabled[1],
                        general$Roch_Population_Disability[1], general$Roch_Population_Nondisabled[1]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_general <- chisq.test(general_table)
print(chisq_result_general)

barplot(general_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(general_table) * 1.4),
        legend = rownames(general_table),
        main = "Count of General Population Difference in Disability Status")

print(chisq_result_general$residuals)
```

We can see that for the general population, the p-value is below 2.2e-16 which is under the p-value limit of 0.05, and can say that there is significant statistical difference in the disability rates between Buffalo and Rochester.  Looking futher at the chi-squared residuals, we can see that the data skews to show Buffalo having a lower disability rate compared to Rochester.

### Population - Between 5 to 17 Years
```{r}
#5 to 17 Years
age_table <- matrix(c(age$Buf_Population_Disability[2], age$Buf_Population_Nondisabled[2],
                        age$Roch_Population_Disability[2], age$Roch_Population_Nondisabled[2]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_age <- chisq.test(age_table)
print(chisq_result_age)

barplot(age_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(age_table) * 1.4),
        legend = rownames(age_table),
        main = "Difference in Disability Status, Age 5 to 17 Years")


chisq_result_age$stdres
```



### Population - Under 5
```{r}
#Under 5
age_table <- matrix(c(age$Buf_Population_Disability[1], age$Buf_Population_Nondisabled[1],
                        age$Roch_Population_Disability[1], age$Roch_Population_Nondisabled[1]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_age <- chisq.test(age_table)
print(chisq_result_age)

barplot(age_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(age_table) * 1.4),
        legend = rownames(age_table),
        main = "Difference in Disability Status, Age Under 5 Years")


chisq_result_age$stdres
```
Here the disability rates for 0 - 5 is significant, with Rochester having lower rates compared to Buffalo for this demographic for the first time in our tests.  However, looking at the data, it seems like the reported population for ages 0 - 5 in Rochester is suspiciously low.  When organizing the primary demographics by rate, we can see the Rochester 0 - 5 age demographic has a rate of 0.002, with the next lowest having a rate of 0.09.

```{r}
general %>%
  arrange(Roch_Disability_Percentage) %>% 
  select(Category, Roch_Disability_Percentage)
  slice_head(n = 3)   
```



### Population - Between 18 and 34 Years
```{r}
#18 to 34 Years
age_table <- matrix(c(age$Buf_Population_Disability[3], age$Buf_Population_Nondisabled[3],
                        age$Roch_Population_Disability[3], age$Roch_Population_Nondisabled[3]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_age <- chisq.test(age_table)
print(chisq_result_age)

barplot(age_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(age_table) * 1.4),
        legend = rownames(age_table),
        main = "Difference in Disability Status, Age 5 to 17 Years")


chisq_result_age$stdres
```

### Population - Two or More Races
```{r}
#Two or More Races
race_table <- matrix(c(race$Buf_Population_Disability[7], race$Buf_Population_Nondisabled[7],
                        race$Roch_Population_Disability[7], race$Roch_Population_Nondisabled[7]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(race_table) * 1.4),
        legend = rownames(race_table),
        main = "Difference in Disability Status, Two or More Races")


chisq_result_race$stdres
```


### Population - White Alone
```{r}
#White
race_table <- matrix(c(race$Buf_Population_Disability[1], race$Buf_Population_Nondisabled[1],
                        race$Roch_Population_Disability[1], race$Roch_Population_Nondisabled[1]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(race_table) * 1.4),
        legend = rownames(race_table),
        main = "Difference in Disability Status, White Alone")


chisq_result_race$stdres
```

### Population - White Alone, Not Hispanic or Latino
```{r}
#White Alone - not Hispanic or Latino
race_table <- matrix(c(race$Buf_Population_Disability[8], race$Buf_Population_Nondisabled[8],
                        race$Roch_Population_Disability[8], race$Roch_Population_Nondisabled[8]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(race_table) * 1.4),
        legend = rownames(race_table),
        main = "Difference in Disability Status, White Alone - Not Hispanic or Latino")


chisq_result_race$stdres
```

### Population - Hispanic or Latino
```{r}
#Hispanic or Latino
race_table <- matrix(c(race$Buf_Population_Disability[9], race$Buf_Population_Nondisabled[9],
                        race$Roch_Population_Disability[9], race$Roch_Population_Nondisabled[9]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        ylim = c(0, max(race_table) * 1.4),
        legend = rownames(race_table),
        main = "Difference in Disability Status, Hispanic or Latino (of any race)")


chisq_result_race$stdres
```

### Population - American Indian and Alaska Native Alone

```{r}
#American Indian and Alaska Native alone

race_table <- matrix(c(race$Buf_Population_Disability[3], race$Buf_Population_Nondisabled[3],
                        race$Roch_Population_Disability[3], race$Roch_Population_Nondisabled[3]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_race <- chisq.test(race_table)
print(chisq_result_race)

barplot(race_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        legend = rownames(race_table),
        ylim = c(0, max(race_table) * 1.4),
        main = "Difference in Disability Status, American Indian and Alaska Native")


chisq_result_race$stdres
```


### Population - Men
```{r}
males_table <- matrix(c(sex$Buf_Population_Disability[1], sex$Buf_Population_Nondisabled[1],
                        sex$Roch_Population_Disability[1], sex$Roch_Population_Nondisabled[1]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_males <- chisq.test(males_table)
print(chisq_result_males)


barplot(males_table,
        beside = TRUE, 
        col = c("lightblue", "pink"),
        legend = rownames(males_table),
        ylim = c(0, max(males_table) * 1.4),
        main = "Difference in Disability Status, Men")


chisq_result_males$stdres

```

### Population - Women
```{r}
females_table <- matrix(c(sex$Buf_Population_Disability[2], sex$Buf_Population_Nondisabled[2],
                        sex$Roch_Population_Disability[2], sex$Roch_Population_Nondisabled[2]),
                      nrow = 2, byrow = TRUE,
                      dimnames = list(c("Buffalo", "Rochester"), c("Disability", "No Disability")))

chisq_result_females <- chisq.test(females_table)
print(chisq_result_females)

barplot(females_table,
        beside = TRUE, 
        col = c("lightblue", "lightpink"),
        legend = rownames(females_table),
        ylim = c(0, max(females_table) * 1.5),
        main = "Difference in Disability Status, Women")


chisq_result_females$stdres

```




# Conclusion








